The Meta-Ontology of AI systems with Human-Level Intelligence

### 2023-02-23 ART_Krzanowski_Polak/Krzanowski_Polak.tex
Editorial corrections added

\begin{artengenv2auth}{Roman Krzanowski, Pawel Polak}
	{The meta-ontology of AI systems with human-level intelligence}
	{The meta-ontology of AI systems with human-level intelligence}
	{The meta-ontology of AI systems with human-level intelligence}
	{\textsuperscript{1}Departments of Philosophy and Mathematics, Carnegie Mellon University\\
		\textsuperscript{2}Copernicus Center for Interdisciplinary Studies, Jagiellonian University}
	{\label{krzanowpolak_start}In this paper, we examine the meta-ontology of AI systems with human-level intelligence, with us denoting such AI systems as AI\textsubscript{E}. Meta-ontology in philosophy is a~discourse centered on ontology, ontological commitment, and the truth condition of ontological theories. We therefore discuss how meta-ontology is conceptualized for AI\textsubscript{E} systems. We posit that the meta-ontology of AI\textsubscript{E~}systems is not concerned with computational representations of reality in the form of structures, data constructs, or computational concepts, while the ontological commitment of AI\textsubscript{E} systems is directed toward what exists in the outside world. Furthermore, the truth condition of the ontology (which is meta-ontological assumption) of AI\textsubscript{E} systems does not require consistency with closed conceptual schema or ontological theories but rather with reality, or in other words, ``what is the world''
	%\label{ref:RNDuTcZpb4G1k}(Smith, 2019, p.57).
	\parencite[][p.57]{smith_promise_2019}. %
	 In addition, the truth condition of AI\textsubscript{E} systems is verified through operational success rather than by coherence with theories. This work builds on ontological postulates about AI systems that were formulated by Brian Caldwell Smith 
	%\label{ref:RND1W3b57kSJx}(2019).
	\parencite*[][]{smith_promise_2019}.%
	}
	{human-level intelligence AI, meta-ontology of AI Paradigm, AI Paradigm, ontology of AI Paradigm, ontological commitment of AI Paradigm, J. Haugeland, B. C. Smith, M. Minsky, H. Dreyfus.}
	{%
%		{\flushright\subbold{Roman Krzanowski}\\\subsubsectit\small{Pontifical University of John Paul II in Krakow}\par}%
%		{\flushright\subbold{Pawel Polak}\\\subsubsectit\small{Pontifical University of John Paul II in Krakow}\par}%
		{\flushright\subbold{Roman Krzanowski}\\\subbold{Pawel Polak}\\\subsubsectit\small{Pontifical University of John Paul II in Krakow}\par}%
	}

@@

In philosophy, ontology is the study of being as it is (i.e., ``what is''), so it is about ``being'' in the most general sense\footnote{The term ``being'' is used in the sense employed by the Ancient Greeks, Parmenides (opposite to The Unbeing), Aristotle (Being qua being), medieval scholars like Aquinas (the study of being qua being)
%\label{ref:RNDFR1vKzYSIx}(Kerr, 2022),
\parencite[][]{kerr_aquinas_2022}, %
 and some modern philosophers such as 
%\label{ref:RNDYFnNv0y86Z}(Jacquette, 2002; Stróżewski, 2004, p.32, or Perzanowski, 2015).
\parencites[][]{jacquette_ontology_2002}[][]{strozewski_ontologia_2004}[or][]{perzanowski_rozprawa_2015}. %
 This term is also sometimes written as Being meaning totality of what exist 
%\label{ref:RNDjNhA4sxPMl}(Kenny, 2012, p.160).
\parencite[][p.160]{kenny_new_2012}. %
 Many modern philosophies, scientists computer engineers infuse this term with many different meanings (see this paper for the examples) obfuscating the original Greek sense of \textit{onto-logia} -- the fundamental study of being, probably as too esoteric (i.e., metaphysical) for their tastes 
%\label{ref:RNDjJRKpH1qcI}(see also Kenny, 2012; Hofweber, 2021).
\parencites[see also][]{kenny_new_2012}[][]{hofweber_logic_2021}.%
}. More specifically, in its purely philosophical meaning, ontology is the study of the foundations of what exists, what is common and most general among it, and what its origins are 
%\label{ref:RNDfYYNHUcznT}(see e.g. Jacquette, 2002; Stróżewski, 2004, p.32)
\parencites[see, e.g.,][]{jacquette_ontology_2002}[][p.32]{strozewski_ontologia_2004} %
 Hereafter, we refer to this concept by a~boldface, capital ``O'' without a~subscript (i.e., \textbf{O}).

@@

\textbf{In computational systems}, ontology can be defined as ``a specific vocabulary (dictionary) used to describe a~certain reality, plus a~set of explicit assumptions regarding the intended meaning of the vocabulary words''
%\label{ref:RNDAcXPc7BmPe}(see Guarino and Giaretta, 1995; Guarino, Oberle and Staab, 2009).
\parencites[see][]{guarino_ontologies_1995}[][]{guarino_what_2009}. %
 In this context, ontologymay also refer to ``a model of the structure of a~system'' 
%\label{ref:RNDi9mTVDxk3A}(Guarino, Oberle and Staab, 2009)
\parencite[][]{guarino_what_2009} %
 or ``a formal, explicit specification of a~shared conceptualization'' 
%\label{ref:RNDAqesK2fTKs}(Studer, Benjamins and Fensel, 1998).
\parencite[][]{studer_knowledge_1998}. %
 We also have computational ontologies, which often called engineering ontologies, that ``are machine-processable structures which represent particular domains of interest'' 
%\label{ref:RNDiOONhTuWRW}(Husáková and Bureš, 2020).
\parencite[][]{husakova_formal_2020}. %
 Ontology may also be used to refer to knowledge-based systems, databases, or AI systems that manage knowledge bases 
%\label{ref:RND7tN9yeZ52M}(see the discussions of Sharman, Kishore and Ramesh, 2007; Staab and Studer, 2009; Garbacz and Oliver Kutz, 2014; Husáková and Bureš, 2020).
\parencites[see the discussions of][]{sharman_ontologies_2007}[][]{staab_handbook_2009}[][]{garbacz_formal_2014}[][]{husakova_formal_2020}.%

\textbf{The ontology of} AI\textsubscript{E} systems, meanwhile, denotes and determines how an AI\textsubscript{E} system represents and reasons about the real world. Itis not concerned with computational representations of reality through structures, theories, data constructs, or computational concepts but rather with how real world objects, properties, and relations are registered by an AI system, as well as how they are recognized and interpreted. In brief, this ontology is solely committed to the real world
%\label{ref:RNDHoOzRM31HN}(in the sense explained by Smith, 2019, p.145),
\parencite[in the sense explained by][p.145]{smith_promise_2019}, %
 with it reflecting the real world\footnote{The term ``real world'', or reality, is understood here as it is understood in Smith 
%\label{ref:RNDdF84GECegW}(2019, pp.xiii, xiv),
\parencite*[][xiv]{smith_promise_2019}, %
 denoting the physical world we live in. Minsky 
%\label{ref:RND166uXGx81U}(1991, p.6)
\parencite*[][p.6]{minsky_logical_1991} %
 refers to this reality as common sense reality. The term may be opposed to ``virtual worlds'', ``imaginary worlds'', ``fantastic worlds'', or other qualified uses of ``words'' denoting worlds as creations of computer systems, artistic expressions, or imaginations. The term ``world'' or ``real world'' may have multiple interpretations that we have no intention to discuss as such a~discussion would be pointless and would not further the main point of the paper. Thus, the reader seeking more detailed explanation of this term should follow the cited references..}, or physical reality, and the AI system's place in this world. More specifically, it is embodied, embedded, extended, and enactive 
%\label{ref:RNDcx3KAaKj2v}(Smith, 2019, p.43),
\parencite[][p.43]{smith_promise_2019}, %
 which are terms that will be explained later in this paper. There is no formal theory to accompany 4E ontology (ontology of AI system that is embodied, embedded, extended, and enactive), so there are no criteria for theoretical truth verification, but verification comes instead from a~confrontation with the real world, which we will discuss later. A~4E ontology is not given in the form of a~set of a~priori relations and objects but rather acquired 
%\label{ref:RNDIdA3C5UrQl}(Smith, 2019)
\parencite[][]{smith_promise_2019} %
 in response to a~dynamically changing reality 
%\label{ref:RNDjzZunSO8Xh}(see Minsky, 1991, as well as; Bołtuć, 2020).
\parencites[see][as well as]{minsky_logical_1991}[][]{boltuc_conscious_2020}. %
 We will refer to this ontology as O\textsubscript{E}.

@@

Schopman
%\label{ref:RNDJc1x1QMFP2}(1986)
\parencite*[][]{schopman_artifical_1986} %
 suggests that AI has not developed a~specific paradigm, claiming that ``…no computational paradigm has yet been produced: there is no single generally accepted way to do AI'' 
%\label{ref:RNDPBgq3pJi3C}(Schopman, 1986, p.6).
\parencite[][p.6]{schopman_artifical_1986}. %
 Čaplinskas 
%\label{ref:RNDMRm8DpWtKy}(1998),
\parencite*[][]{caplinskas_ai_1998}, %
 however, defines three AI paradigms: the behaviorist paradigm, the agent paradigm, and the artificial life paradigm. Norvig 
%\label{ref:RNDlOHT6rHNA2}(1992; 2002),
\parencites*[][]{norvig_paradigms_1992}[][]{norvig_retrospective_2002}, %
 meanwhile, associates the term with Lisp programming to express the paradigm of Artificial Intelligence Programming as being equivalent to a~programming approach. Next, Cristianini (2014) distinguishes four AI paradigms: data-driven AI, statistical AI, knowledge-driven AI, and reasoning-and-search-based AI. Without much explanation as to why, Leary 
%\label{ref:RNDMgW31BP6Zx}(2017)
\parencite*[][]{leary_googles_2017} %
 claims that the new Google AI paradigm is machine learning, while in a~blog post titled ``The AI Paradigm Shift,'' Richardson 
%\label{ref:RNDt2zoIcErAM}(2018)
\parencite*[][]{richardson_ai_2018} %
 refers broadly to the AI paradigm as an approach to engineering AI systems, such as deep learning (DL), machine learning (ML), natural language processing (NLP), and robotics. 
%\label{ref:RNDcu5g1TESmD}(Hernández-Orallo et al., 2020, p.2522),
\parencite[][p.2522]{hernandez-orallo_ai_2020}, %
 meanwhile, claims that the concepts of AI paradigms have been used to denote ``broad families of technical or conceptual approaches: ‘symbolic' vs ‘connectionist', reasoning vs learning, expert systems vsagents.''

Much like Richardson, Romero
%\label{ref:RNDPl3RwoIA7v}(2021)
\parencite*[][]{romero_unpopular_2021} %
 refers to deep learning and machine learning as AI paradigms, while Yalçın 
%\label{ref:RNDmyfcw0kayw}(2021)
\parencite*[][]{osaba_artificial_2021} %
 refers to symbolic (i.e., a~human-readable, symbolic representation of problems, logic, and search) and sub-symbolic (i.e., an implicit representation derived from experience-based learning with no symbolic representation of rules and properties) representations as AI paradigms. Villar et al. 
%\label{ref:RNDY4KjMoWStM}(2021)
\parencite*[][]{xu_artificial_2021} %
 hint at relating the AI paradigm to versions of machine-learning and deep-learning methods. Meanwhile, Luhach Kumar and El\c{c}i Atilla 
%\label{ref:RND1BjB3dukU7}(2021)
\parencite*[][]{luhach_artificial_2021} %
 in their book use the term ``paradigm'' in several places but in different contexts, with its meaning being variably associated with programming, the applications of AI to smart computational cyberspaces, or execution paradigms associated with computer hardware. Next, Joseph Makokha 
%\label{ref:RNDS2MHw4L6Mk}(2021)
\parencite*[][]{makokha_artificial_2021} %
 distinguishes two AI paradigms, namely an AI-based one for rule-following methods and another one based on artificial neural network constructs. The term ``AI paradigm'' is often used to simply denote a~method for AI learning or knowledge acquisition, and this is how Yonguin Xu 
%\label{ref:RNDGRGBuNdPMm}(Xu et al., 2021)
\parencite[][]{xu_artificial_2021} %
 uses this term to denote~deep-learning approaches, such as supervised, unsupervised, and reinforced learning.

Several studies have posited that the current AI systems use two broad, conceptual constructs, namely the symbolic and the sub-symbolic
%\label{ref:RNDiWrk1ACHd1}(e.g. Searle, 1998; Harvey, 2013; Neapolitan and Jiang, 2018; Mitchell, 2019; Smith, 2019; Cole, 2020; Russell and Norvig, 2020; Wooldridge, 2021).
\parencites[e.g.,][]{searle_mind_1998}[][]{harvey_perspectives_2013}[][]{neapolitan_artificial_2018}[][]{mitchell_artificial_2019}[][]{smith_promise_2019}[][]{cole_chinese}[][]{russell_artificial_2020}[][]{wooldridge_road_2021}. %
 Thus, the symbolic paradigm reflects symbolic representations of a~priori defined concepts that may be implemented in various programming environments, while the sub-symbolic paradigm relates to less clearly defined concepts, such as the multi-dimensional probability weights on the connections within an artificial neural network. An approach under the sub-symbolic paradigm would therefore be implemented using one of the various machine learning (ML) technologies. These two paradigms can also be fused into a~neuro-symbolic paradigm 
%\label{ref:RNDTv0qtZz01g}(e.g. Bader and Hitzler, 2005; Garcez et al., 2019; Garcez and Lamb, 2020; Kautz, 2022)
\parencites[e.g.,][]{bader_dimensions_2005}[][]{garcez_neural-symbolic_2019}[][]{garcez_neurosymbolic_2020}[][]{kautz_third_2022} %
 that combines symbolic and sub-symbolic elements. As pointed out earlier, the term ``AI paradigm'' does not point to a~single software or hardware solution 
%\label{ref:RNDzI0NRHZm9F}(see Minsky, 1991; Russell and Norvig, 2020)
\parencites[see][]{minsky_logical_1991}[][]{russell_artificial_2020}%

@@

\section*{The meta-ontology of AI\textsubscript{E} systems}
We have concluded that in philosophy, meta-ontology is the study of what ontology is all about. In other words, it is a~study of, or about, ontology, as well as what it investigates; after Berto and Plebani ``‘meta-X' as the inquiry on the central concepts and procedures of discipline X''where ``X'' refers to ontology in this case
%\label{ref:RNDc30wpAwAYr}(Berto and Plebani, 2015, p.13).
\parencite[][p.13]{berto_ontology_2015}. %
 We also concluded that ontological commitment and truth conditions represent the key concepts of meta-ontology, and these terms are critical for differentiating between ontology and meta-ontology.

@@

%(\textbf{T1) The ontological commitment of AI}\textbf{\textsubscript{E}}~\textbf{is to the real world,}\footnote{The term explained earlier in the text.}
%\textbf{the world of a~human agent}:
\paragraph[(T1) The ontological commitment]{(T1) The ontological commitment of AI\textsubscript{E} is to the real world,\footnote{The term explained earlier in the text.} world of a~human agent.}%
 The world for AI\textsubscript{E} is the same reality that a~human actor would exist in. We could say that AI\textsubscript{E} ontology is a~partial ontology as opposed to one that covers the entire world, so it is about a~state of affairs, by which we mean a~local, temporal, dynamic (as described by Minsky) reality of the everyday world. This partial ontology does not attempt to create a~comprehensive ontology of existence but rather account for what exists, together with the state of affairs,\footnote{The term ``state of affairs'' is used in the sense employed by Jacquette
%\label{ref:RNDlvlS6tqNBF}(2002).
\parencite*[][]{jacquette_ontology_2002}.%
} in the part of the actual world that is relevant to an AI\textsubscript{E} agent, we may say, agent-relevant ontology.

@@

The truth condition of the AI\textsubscript{E} paradigm shares some similarities with the truth condition of philosophy, which states that ``theory is ontologically committed to an object only if that object occurs in all the ontologies of that theory''
%\label{ref:RNDe7Qk8dHvAr}(Gibson, 2009, p.631).
\parencite[][p.631]{gibson_ontological_2009}. %
 Or, in and alternative formulation by Rayo 
%\label{ref:RNDez2d7KVZBH}(Rayo, 2007)
\parencite[][]{rayo_ontological_2007} %
 rephrased as follows: ``for a~sentence to carry commitment to Fs is for the sentence's truth to demand of the world that it contain Fs.'' However, the truth condition of AI\textsubscript{E} systems requires AI systems to ``deal with reality as it actually is---not in the way our language represents it as being'' 
%\label{ref:RNDPlQIuEJgnp}(Smith, 2019, p.34),
\parencite[][p.34]{smith_promise_2019}, %
 i.e., the truth condition of AI\textsubscript{E} systems does not have to satisfy any sentences; understood as the ontological claims expressed in some form of language.

>>>>

\begin{artengenv2auth}{Roman Krzanowski, Pawel Polak}
	{The meta-ontology of AI systems with human-level intelligence}
	{The meta-ontology of AI systems with human-level intelligence}
	{The meta-ontology of AI systems with human-level intelligence}
	{\textsuperscript{1}Departments of Philosophy and Mathematics, Carnegie Mellon University\\
		\textsuperscript{2}Copernicus Center for Interdisciplinary Studies, Jagiellonian University}
	{\label{krzanowpolak_start}In this paper, we examine the meta-ontology of AI systems with human-level intelligence, with us denoting such AI systems as AI\textsubscript{E}. Meta-ontology in philosophy is a~discourse centered on ontology, ontological commitment, and the truth condition of ontological theories. We therefore discuss how meta-ontology is conceptualized for AI\textsubscript{E} systems. We posit that the meta-ontology of AI\textsubscript{E~}systems is not concerned with computational representations of reality in the form of structures, data constructs, or computational concepts, while the ontological commitment of AI\textsubscript{E} systems is directed toward what exists in the outside world. Furthermore, the truth condition of the ontology (which is meta-ontological assumption) of AI\textsubscript{E} systems does not require consistency with closed conceptual schema or ontological theories but rather with reality, or in other words, ``what is the world''
	%\label{ref:RNDuTcZpb4G1k}(Smith, 2019, p.57).
	\parencite[][p.57]{smith_promise_2019}. %
	 In addition, the truth condition of AI\textsubscript{E} systems is verified through operational success rather than by coherence with theories. This work builds on ontological postulates about AI systems that were formulated by Brian Caldwell Smith 
	%\label{ref:RND1W3b57kSJx}(2019).
	\parencite*[][]{smith_promise_2019}.%
	}
	{human-level intelligence AI, meta-ontology of AI Paradigm, AI Paradigm, ontology of AI Paradigm, ontological commitment of AI Paradigm, John Haugeland, Brian Cantwell Smith, Marvin Minsky, Hubert Dreyfus.}
	{%
%		{\flushright\subbold{Roman Krzanowski}\\\subsubsectit\small{Pontifical University of John Paul II in Krakow}\par}%
%		{\flushright\subbold{Pawel Polak}\\\subsubsectit\small{Pontifical University of John Paul II in Krakow}\par}%
		{\flushright\subbold{Roman Krzanowski}\\\subbold{Pawel Polak}\\\subsubsectit\small{Pontifical University of John Paul II in Kraków}\par}%
	}

@@

In philosophy, ontology is the study of being as it is (i.e., ``what is''), so it is about ``being'' in the most general sense.\footnote{The term ``being'' is used in the sense employed by the Ancient Greeks, Parmenides (opposite to The Unbeing), Aristotle (Being qua being), medieval scholars like Aquinas (the study of being qua being)
%\label{ref:RNDFR1vKzYSIx}(Kerr, 2022),
\parencite[][]{kerr_aquinas_2022}, %
 and some modern philosophers such as 
%\label{ref:RNDYFnNv0y86Z}(Jacquette, 2002; Stróżewski, 2004, p.32, or Perzanowski, 2015).
\parencites[][]{jacquette_ontology_2002}[][]{strozewski_ontologia_2004}[or][]{perzanowski_rozprawa_2015}. %
 This term is also sometimes written as Being meaning totality of what exist 
%\label{ref:RNDjNhA4sxPMl}(Kenny, 2012, p.160).
\parencite[][p.160]{kenny_new_2012}. %
 Many modern philosophies, scientists computer engineers infuse this term with many different meanings (see this paper for the examples) obfuscating the original Greek sense of \textit{onto-logia} -- the fundamental study of being, probably as too esoteric (i.e., metaphysical) for their tastes 
%\label{ref:RNDjJRKpH1qcI}(see also Kenny, 2012; Hofweber, 2021).
\parencites[see also][]{kenny_new_2012}[][]{hofweber_logic_2021}.%
} More specifically, in its purely philosophical meaning, ontology is the study of the foundations of what exists, what is common and most general among it, and what its origins are 
%\label{ref:RNDfYYNHUcznT}(see e.g. Jacquette, 2002; Stróżewski, 2004, p.32)
\parencites[see, e.g.,][]{jacquette_ontology_2002}[][p.32]{strozewski_ontologia_2004}. %
 Hereafter, we refer to this concept by a~boldface, capital ``O'' without a~subscript (i.e., \textbf{O}).

@@

In computational systems, ontology can be defined as ``a specific vocabulary (dictionary) used to describe a~certain reality, plus a~set of explicit assumptions regarding the intended meaning of the vocabulary words''
%\label{ref:RNDAcXPc7BmPe}(see Guarino and Giaretta, 1995; Guarino, Oberle and Staab, 2009).
\parencites[see][]{guarino_ontologies_1995}[][]{guarino_what_2009}. %
 In this context, ontologymay also refer to ``a model of the structure of a~system'' 
%\label{ref:RNDi9mTVDxk3A}(Guarino, Oberle and Staab, 2009)
\parencite[][]{guarino_what_2009} %
 or ``a formal, explicit specification of a~shared conceptualization'' 
%\label{ref:RNDAqesK2fTKs}(Studer, Benjamins and Fensel, 1998).
\parencite[][]{studer_knowledge_1998}. %
 We also have computational ontologies, which often called engineering ontologies, that ``are machine-processable structures which represent particular domains of interest'' 
%\label{ref:RNDiOONhTuWRW}(Husáková and Bureš, 2020).
\parencite[][]{husakova_formal_2020}. %
 Ontology may also be used to refer to knowledge-based systems, databases, or AI systems that manage knowledge bases 
%\label{ref:RND7tN9yeZ52M}(see the discussions of Sharman, Kishore and Ramesh, 2007; Staab and Studer, 2009; Garbacz and Oliver Kutz, 2014; Husáková and Bureš, 2020).
\parencites[see the discussions of][]{sharman_ontologies_2007}[][]{staab_handbook_2009}[][]{garbacz_formal_2014}[][]{husakova_formal_2020}.%

The ontology of AI\textsubscript{E} systems, meanwhile, denotes and determines how an AI\textsubscript{E} system represents and reasons about the real world. Itis not concerned with computational representations of reality through structures, theories, data constructs, or computational concepts but rather with how real world objects, properties, and relations are registered by an AI system, as well as how they are recognized and interpreted. In brief, this ontology is solely committed to the real world
%\label{ref:RNDHoOzRM31HN}(in the sense explained by Smith, 2019, p.145),
\parencite[in the sense explained by][p.145]{smith_promise_2019}, %
 with it reflecting the real world\footnote{The term ``real world'', or reality, is understood here as it is understood in Smith 
%\label{ref:RNDdF84GECegW}(2019, pp.xiii, xiv),
\parencite*[][xiv]{smith_promise_2019}, %
 denoting the physical world we live in. Minsky 
%\label{ref:RND166uXGx81U}(1991, p.6)
\parencite*[][p.6]{minsky_logical_1991} %
 refers to this reality as common sense reality. The term may be opposed to ``virtual worlds'', ``imaginary worlds'', ``fantastic worlds'', or other qualified uses of ``words'' denoting worlds as creations of computer systems, artistic expressions, or imaginations. The term ``world'' or ``real world'' may have multiple interpretations that we have no intention to discuss as such a~discussion would be pointless and would not further the main point of the paper. Thus, the reader seeking more detailed explanation of this term should follow the cited references..}, or physical reality, and the AI system's place in this world. More specifically, it is embodied, embedded, extended, and enactive 
%\label{ref:RNDcx3KAaKj2v}(Smith, 2019, p.43),
\parencite[][p.43]{smith_promise_2019}, %
 which are terms that will be explained later in this paper. There is no formal theory to accompany 4E ontology (ontology of AI system that is embodied, embedded, extended, and enactive), so there are no criteria for theoretical truth verification, but verification comes instead from a~confrontation with the real world, which we will discuss later. A~4E ontology is not given in the form of a~set of a~priori relations and objects but rather acquired 
%\label{ref:RNDIdA3C5UrQl}(Smith, 2019)
\parencite[][]{smith_promise_2019} %
 in response to a~dynamically changing reality 
%\label{ref:RNDjzZunSO8Xh}(see Minsky, 1991, as well as; Bołtuć, 2020).
\parencites[see][as well as]{minsky_logical_1991}[][]{boltuc_conscious_2020}. %
 We will refer to this ontology as O\textsubscript{E}.

@@

Schopman
%\label{ref:RNDJc1x1QMFP2}(1986)
\parencite*[][]{schopman_artifical_1986} %
 suggests that AI has not developed a~specific paradigm, claiming that ``[\ldots] no computational paradigm has yet been produced: there is no single generally accepted way to do AI'' 
%\label{ref:RNDPBgq3pJi3C}(Schopman, 1986, p.6).
\parencite[][p.6]{schopman_artifical_1986}. %
 Čaplinskas 
%\label{ref:RNDMRm8DpWtKy}(1998),
\parencite*[][]{caplinskas_ai_1998}, %
 however, defines three AI paradigms: the behaviorist paradigm, the agent paradigm, and the artificial life paradigm. Norvig 
%\label{ref:RNDlOHT6rHNA2}(1992; 2002),
\parencites*[][]{norvig_paradigms_1992}[][]{norvig_retrospective_2002}, %
 meanwhile, associates the term with Lisp programming to express the paradigm of Artificial Intelligence Programming as being equivalent to a~programming approach. Next, Cristianini (2014) distinguishes four AI paradigms: data-driven AI, statistical AI, knowledge-driven AI, and reasoning-and-search-based AI. Without much explanation as to why, Leary 
%\label{ref:RNDMgW31BP6Zx}(2017)
\parencite*[][]{leary_googles_2017} %
 claims that the new Google AI paradigm is machine learning, while in a~blog post titled ``The AI Paradigm Shift,'' Richardson 
%\label{ref:RNDt2zoIcErAM}(2018)
\parencite*[][]{richardson_ai_2018} %
 refers broadly to the AI paradigm as an approach to engineering AI systems, such as deep learning (DL), machine learning (ML), natural language processing (NLP), and robotics. 
%\label{ref:RNDcu5g1TESmD}(Hernández-Orallo et al., 2020, p.2522),
\parencite[][p.2522]{hernandez-orallo_ai_2020}, %
 meanwhile, claims that the concepts of AI paradigms have been used to denote ``broad families of technical or conceptual approaches: ‘symbolic' vs ‘connectionist', reasoning vs learning, expert systems vs agents.''

Much like Richardson, Romero
%\label{ref:RNDPl3RwoIA7v}(2021)
\parencite*[][]{romero_unpopular_2021} %
 refers to deep learning and machine learning as AI paradigms, while Yalçın 
%\label{ref:RNDmyfcw0kayw}(2021)
\parencite*[][]{osaba_artificial_2021} %
 refers to symbolic (i.e., a~human-readable, symbolic representation of problems, logic, and search) and sub-symbolic (i.e., an implicit representation derived from experience-based learning with no symbolic representation of rules and properties) representations as AI paradigms. Villar et al. 
%\label{ref:RNDY4KjMoWStM}(2021)
\parencite*[][]{xu_artificial_2021} %
 hint at relating the AI paradigm to versions of machine-learning and deep-learning methods. Meanwhile, Luhach Kumar and El\c{c}i Atilla 
%\label{ref:RND1BjB3dukU7}(2021)
\parencite*[][]{luhach_artificial_2021} %
 in their book use the term ``paradigm'' in several places but in different contexts, with its meaning being variably associated with programming, the applications of AI to smart computational cyberspaces, or execution paradigms associated with computer hardware. Next, Joseph Makokha 
%\label{ref:RNDS2MHw4L6Mk}(2021)
\parencite*[][]{makokha_artificial_2021} %
 distinguishes two AI paradigms, namely an AI-based one for rule-following methods and another one based on artificial neural network constructs. The term ``AI paradigm'' is often used to simply denote a~method for AI learning or knowledge acquisition, and this is how Yonguin Xu et al. 
%\label{ref:RNDGRGBuNdPMm}(Xu et al., 2021)
\parencite*[][]{xu_artificial_2021} %
 use this term to denote~deep-learning approaches, such as supervised, unsupervised, and reinforced learning.

Several studies have posited that the current AI systems use two broad, conceptual constructs, namely the symbolic and the sub-symbolic
%\label{ref:RNDiWrk1ACHd1}(e.g. Searle, 1998; Harvey, 2013; Neapolitan and Jiang, 2018; Mitchell, 2019; Smith, 2019; Cole, 2020; Russell and Norvig, 2020; Wooldridge, 2021).
\parencites[e.g.,][]{searle_mind_1998}[][]{harvey_perspectives_2013}[][]{neapolitan_artificial_2018}[][]{mitchell_artificial_2019}[][]{smith_promise_2019}[][]{cole_chinese}[][]{russell_artificial_2020}[][]{wooldridge_road_2021}. %
 Thus, the symbolic paradigm reflects symbolic representations of a~priori defined concepts that may be implemented in various programming environments, while the sub-symbolic paradigm relates to less clearly defined concepts, such as the multi-dimensional probability weights on the connections within an artificial neural network. An approach under the sub-symbolic paradigm would therefore be implemented using one of the various machine learning (ML) technologies. These two paradigms can also be fused into a~neuro-symbolic paradigm 
%\label{ref:RNDTv0qtZz01g}(e.g. Bader and Hitzler, 2005; Garcez et al., 2019; Garcez and Lamb, 2020; Kautz, 2022)
\parencites[e.g.,][]{bader_dimensions_2005}[][]{garcez_neural-symbolic_2019}[][]{garcez_neurosymbolic_2020}[][]{kautz_third_2022} %
 that combines symbolic and sub-symbolic elements. As pointed out earlier, the term ``AI paradigm'' does not point to a~single software or hardware solution 
%\label{ref:RNDzI0NRHZm9F}(see Minsky, 1991; Russell and Norvig, 2020)
\parencites[see][]{minsky_logical_1991}[][]{russell_artificial_2020}.%

@@

\section*{The meta-ontology of AI\textsubscript{E} systems}
We have concluded that in philosophy, meta-ontology is the study of what ontology is all about. In other words, it is a~study of, or about, ontology, as well as what it investigates; after Berto and Plebani ``‘meta-X' as the inquiry on the central concepts and procedures of discipline X'', where ``X'' refers to ontology in this case
%\label{ref:RNDc30wpAwAYr}(Berto and Plebani, 2015, p.13).
\parencite[][p.13]{berto_ontology_2015}. %
 We also concluded that ontological commitment and truth conditions represent the key concepts of meta-ontology, and these terms are critical for differentiating between ontology and meta-ontology.

@@

%(\textbf{T1) The ontological commitment of AI}\textbf{\textsubscript{E}}~\textbf{is to the real world,}\footnote{The term explained earlier in the text.}
%\textbf{the world of a~human agent}:
\paragraph[(T1) The ontological commitment]{(T1) The ontological commitment of AI\textsubscript{E} is to the real world,\footnote{The term explained earlier in the text. See footnote 7.} world of a~human agent.}%
 The world for AI\textsubscript{E} is the same reality that a~human actor would exist in. We could say that AI\textsubscript{E} ontology is a~partial ontology as opposed to one that covers the entire world, so it is about a~state of affairs, by which we mean a~local, temporal, dynamic (as described by Minsky) reality of the everyday world. This partial ontology does not attempt to create a~comprehensive ontology of existence but rather account for what exists, together with the state of affairs,\footnote{The term ``state of affairs'' is used in the sense employed by Jacquette
%\label{ref:RNDlvlS6tqNBF}(2002).
\parencite*[][]{jacquette_ontology_2002}.%
} in the part of the actual world that is relevant to an AI\textsubscript{E} agent, we may say, agent-relevant ontology.

@@

The truth condition of the AI\textsubscript{E} paradigm shares some similarities with the truth condition of philosophy, which states that ``theory is ontologically committed to an object only if that object occurs in all the ontologies of that theory''
%\label{ref:RNDe7Qk8dHvAr}(Gibson, 2009, p.631).
\parencite[][p.631]{gibson_ontological_2009}. %
 Or, in and alternative formulation by Rayo 
%\label{ref:RNDez2d7KVZBH}(Rayo, 2007)
\parencite*[][]{rayo_ontological_2007} %
 rephrased as follows: ``for a~sentence to carry commitment to Fs is for the sentence's truth to demand of the world that it contain Fs.'' However, the truth condition of AI\textsubscript{E} systems requires AI systems to ``deal with reality as it actually is---not in the way our language represents it as being'' 
%\label{ref:RNDPlQIuEJgnp}(Smith, 2019, p.34),
\parencite[][p.34]{smith_promise_2019}, %
 i.e., the truth condition of AI\textsubscript{E} systems does not have to satisfy any sentences; understood as the ontological claims expressed in some form of language.

%(\textbf{T3) The truth condition of AI}\textbf{\textsubscript{E}} \textbf{systems} \textbf{is verified through operational success}.
\paragraph{(T3) The truth condition of AI\textsubscript{E} systems is verified through operational success.}
The truth condition of \textbf{AI}\textbf{\textsubscript{E}} \textbf{systems} is not concerned with computational representations of reality in the form of structures, data constructs, or computational concepts. Objects that are registered in \textbf{AI}\textbf{\textsubscript{E}} \textbf{systems} follow~constitutive regularities and norms, but these are not known beforehand but instead derived and learned from the real world as the basis of being
%\label{ref:RNDlXo6W1ixLP}(Smith, 2019, p.103).
\parencite[][p.103]{smith_promise_2019}. %
 There is no theory to go with it, so there is no criterion for truth verification that references a~theory, at least if we accept that this statement is not a~theory in itself. As Baker says in an article about the metaphysics of ordinary things, ``the ultimate test of a~metaphysical theory is\ldots pragmatic'' 
%\label{ref:RNDPeMuxppH4c}(Baker, 2007, p.11).
\parencite[][p.11]{baker_metaphysics_2007}.%
\footnote{In Baker's ontology, the pragmatic mode of verification has nothing to do with pragmatic theories of truth in philosophy.} In other words, the truth condition of AI\textsubscript{E} systems is verified pragmatically, i.e., through operational success, because they are solely committed to the world and their actions within it 
%\label{ref:RNDPFPJ38CJSH}(Smith, 2019, p.145).
\parencite[][p.145]{smith_promise_2019}. %
 The precise meaning of operational success of engineering (including AI\textsubscript{E} systems) or natural systems (living organisms) depend on the specific system the term ``operational success'' is applied to. In biological systems operational success (mostly) means survival and reproduction. In artificial systems operational success means fulfilling design objectives. It is not always obvious what is operational success even in engineering systems. For factory robots operational success is a~well-defined task -- like proper welding of a~pin or similar. For autonomous vehicles operational success means (among other things) collision avoidance. For AI\textsubscript{E} systems operational success is proper response/decision to situations. Of course operational success is much harder to evaluate in some cases than a~welding of a~pin; like it is in a~case of the notorious trolley problem 
%\label{ref:RNDFJRT1yhI7Z}(see e.g. Cathcart, 2013).
\parencite[see e.g.,][]{cathcart_trolley_2013}. %
 ``Operational success'' of the trolley problem is a~subject of endless debates between engineers, philosophers and enthusiasts of AI probably with a~limited chance of success as these groups talk past each other; philosophers see the trolley problem as ethical problems, engineers as engineering problem, and enthusiasts of AI are too emotionally engaged to be rational. As we mentioned earlier, registered/recognized objects in \textbf{AI}\textbf{\textsubscript{E}} \textbf{ontology} create regularities and norms, but rather than being known a~priori, they are derived and learned from the real world, and this provides the grounding for AI\textsubscript{E} ontology.
